---
title: "BCFISHPASS Gradient Estimate Validation"
author: "S Lingard"
date: "`r Sys.Date()`"
output: pdf_document
---
<style>
p.caption {
  font-size: 0.8em;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This document describes the methods applied to compare stream gradients estimated from the Freshwater Atlas of British Columbia, with an existing database of field surveyed stream gradients. The purpose of this exercise is to determine the most accurate method for estimating stream gradient using the available spatial data for the Province of British Columbia. The Freshwater Atlas (FWA) was derived from the Province of British Columbia's 1:20,000 scale topographic base maps and a 25 m Digital Elevation Model (DEM). There is an inherent margin of error in the Freshwater Atlas, and determining this scale of this error in estimating gradient is important to users of this database who may wish to use it for fisheries research or conservation planning purposes. 

All species have a specific niche in which their fitness is maximized. For fish, stream gradient is often a key factor dictating space use because this factor affects the availability of food as well as water velocity which together influence energetic demand. Gradient also dictates the accessibility of habitats.

Two methods have been proposed to estimate stream gradient from the FWA for use in the BCFISHPASS model. The FWA contains irregularly spaced vertices along each line feature representing a water body. At each of these vertices, elevation has been estimated using a 25m DEM. The first gradient estimation method simply calculates the percent slope (gradient) between each vertex and the adjacent vertex moving from the mouth of the tributary upstream. Another method has been developed to estimate gradient for 100 m segments upstream of each vertex using interpolated elevations for the end of the segment based on elevations of nearby vertices. In areas where vertices are less than 100 m apart, this second method creates overlapping segments. 

Forestry licencees are required to measure stream characteristics and gradients to determine if streams are fish bearing or not. In British Columbia, a database of stream measurements has been made publicly available through the BC Data Catelogue (https://catalogue.data.gov.bc.ca/dataset/stream-inventory-sample-sites). The stream inventory database contains geo-located measurements of stream gradient, the length over which gradient is measured, as well as other variables such as substrate and channel width. The stream measurements have been snapped to FWA water bodies using an un-described process. Gradient is often measured in the field using clinometers and survey rods which are simple tools that rely on human vision. Thus, measurements over long distances in this database are likely of low precision.

There are sources of error in both databases to be used in this exercise. In the FWA, there is likely error in both the location of line features representing streams and vertex elevation. There are also sources of error in the stream survey data due to the snapping of measurements to the closest stream, and un-quantifiable measurement error in the gradient estimates themselves. Given the multiple sources of error in each database, we do not expect a perfect relationship between model estimated and field surveyed gradients. We are seeking to determine which of the two methods best reflect field survey data. Although there are major issues with the FWA, in the absence of province wide LiDAR this database is the best available spatial tool for province wide modeling of freshwater ecosystems and potential stressors effects on these systems.

# Methods: 

The model derived gradient measurements were matched to the nearest stream survey record in the following ways. In the vertex-to-vertex model, the gradient for the segment containing the observation point as well as the adjacent upstream and downstream segments were matched to a specific survey record. For the 100 m estimates, the vertex closest to the measurement was used as the central segment, and then estimates from the vertices upstream and downstream of the closest were also matched to the record. 

The first step in the exercise was to explore the data, both the shape of the distributions as well as the minimum and maximum values for both gradient and the length over which gradient was measured (stream surveys, and vertex-to-vertex model). 

```{r load data and libraries, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(lubridate)
library(ggpubr)

here()

dat <- read.csv(here("data","fiss_sample_sites_gradient.csv"))

# vertex to vertex measures occur on average over < 100 m while the surveys mostly occur over 100 m. I will add some columns
dat <- dat %>% 
  rowwise() %>%
  mutate(gradient=abs(gradient),
         gradient100m=abs(gradient100m),
         surveyed_gradient=abs(surveyed_gradient),
    gradient_avg=sum(gradient_upstr, gradient, gradient_dnstr)/3,
         gradient_avg_length=sum(gradient_upstr_length, gradient_length, gradient_dnstr_length))

```

```{r exploratory plots, echo=FALSE, fig.cap="Distribution of values for segment lengths in surveyed data and FWA model derived gradients using vertex to vertex measures for segments containing the measurment point, and an average of gradient across the segment containing the measurment, as well as upstream and downstream segments. Data has been filtered to include only stream orders 1-4 because segment lengths for larger rivers were very long in and skew the data further", fig.width=4.5, fig.height=5.5, fig.align="left"}

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

dat.long <- dat %>%
  select(surveyed_gradient, surveyed_length, gradient, gradient_length, gradient100m, gradient_avg, gradient_avg_length, stream_order)%>%
  gather(., key="variable", value="value", -stream_order)

var.names <- c(
  'gradient_length'="Nearest vertex to vertex seg length",
  'gradient_avg_length'="length of nearest and adjacent vertex-to-vertex segs",
  'surveyed_length'="surveyed length"
)

dat.long %>%  
  filter(stream_order < 5)%>%
  filter(variable %in% c("surveyed_length","gradient_length", "gradient_avg_length") & value < 2000)%>%
  mutate(variable=factor(variable, levels=c("surveyed_length","gradient_length", "gradient_avg_length")))%>%
  ggplot(., aes(x=value))+
  geom_histogram(binwidth=25)+
  facet_wrap(~variable, ncol=1,labeller = as_labeller(var.names))+
 theme_classic()+
  theme(axis.text = element_text(size=12),
        axis.title= element_text(size=14),
        plot.margin = margin(0, 20, 0, 0))+
  scale_x_continuous(labels = scales::comma, expand = c(0,0))+
  scale_y_continuous(labels=scales::comma,expand = c(0,0))+
  labs(y="Count", x="Segment Length (m)")
 
  
  
```


```{r fig.2, fig.cap="Distribution of field survey and model derived gradients.", echo=FALSE, fig.width=4.5, fig.height=5.5, fig.align="left"}


var.names2 <- c(
  'gradient'="gradient of nearest vertex to vertex segment",
  'gradient_avg'="gradient (avg) of nearest, and adjacent v-to-v segments",
  'surveyed_gradient'="surveyed gradient"
)

dat.long %>%
  filter(variable %in% c("surveyed_gradient","gradient", "gradient_avg") & value < 200)%>%
  mutate(variable=factor(variable, levels=c("surveyed_gradient","gradient", "gradient_avg")))%>%
  ggplot(., aes(x=value))+
  geom_histogram(binwidth=5)+
  facet_wrap(~variable, ncol=1,labeller = as_labeller(var.names2))+
  theme_classic()+
  theme(axis.text = element_text(size=12),
        axis.title= element_text(size=14))+
  scale_x_continuous(labels = scales::comma, expand = c(0,0))+
  scale_y_continuous(labels=scales::comma, expand = c(0,0))+
  labs(y="Count", x="Gradient (%)")
  

```

The exploratory plots indicate there are likely errors in the field survey database. For example, it was not uncommon for gradient to have been measured over hundreds or thousands of meters (Figure 1). These measurements are unlikely to be precise and may also represent data entry errors. Given that the Provincial guidance documents specify that forestry licencees should measure gradient over 100 m, all measurements taken over > 100 m were filtered out of the dataset. This decreased the size of the dataset from 61,831 to 25,776. Next, to ensure comparisons were the same across the two modeled gradients, any records that had NA for either vertex-to-vertex gradient, or gradient 100 m were also removed. This resulted in a final dataset of 20,399 observations.

The vertex-to-vertex gradient segments were generally much shorter than the field surveyed segments (Figure 1). To ensure better comparability between these measures and the survey segements, two more values were generated for each observation. First, the average gradient of the nearest vertex, as well as the upstream and downstream segments was calculated. Second, the total length of this combined gradient segment was also calculated. These new variables are called gradient_avg and gradient_avg_length, respectively.

Due to the skew of the data and the presence of outliers, the Spearman rank correlation coefficient was chosen for exploration of the relationship between the surveyed gradients and modeled gradients. The higher the Spearman's rho (R in plots below) the stronger the association. The data were split by stream order for this step of the exercise as there is likely to be more variability in gradient in smaller order streams that large rivers. Few survey records were available for stream orders greater than 5 (< 100 per order), so these categories were excluded from the correlation exercise. 


```{r Data Trimming, echo=FALSE}

trimmed.dat <- dat %>%
  filter(surveyed_length <= 100)%>%
  drop_na(gradient, gradient100m, surveyed_gradient)%>%
  mutate(surveyed_gradient=abs(surveyed_gradient),
         gradient100m=abs(gradient100m))
```

# Results:

The 100m interpolated method for estimating gradient appears to produce values that are marginally more reflective of the field surveyed gradients. However, there is still substantial variability between field measured and modeled derived estimates of gradient. In some cases, there was 20-30% difference in modeled versus measured gradients and in both directions. While 70% correlation would be seen as high in most ecological contexts, small changes in gradient (2%) can create enormous differences in stream reach dynamics. Further, the threshold at which gradient becomes a barrier to salmon is thought to occur between 15-25% slope, depending on species, and which expert is queried. Based on the scale of variability between measured and model derived gradient estimates here, field or literature derived values for accessibility thresholds are unlikely to be useful for projects using FWA derived estimates of gradient. However, there may be alternative ways of parsing or analyzing this data that would give better confidence the reliability of modeled gradients to reality. 

```{r Spearman correlation with interpolated method, fig.cap="Correlation between field surveyed and 100 m interpolated method using Spearman's correltation coefficient (R)", fig.width=8, fig.height=7, echo=FALSE, fig.align="left"}

trimmed.dat %>%
  filter(stream_order < 6)%>%
  ggplot(., aes(x=surveyed_gradient, y=gradient100m))+
  geom_point()+
  facet_wrap(~stream_order, scales="free")+
  stat_cor(aes(label = after_stat(r.label)), method="spearman")+
  labs(x="Field Surveyed Gradient (%)", 
       y="Gradient estimated 100m upstream of each vertex (%)")+
  theme_classic()+
  theme(axis.text=element_text(size=12),
        strip.text = element_text(size=14),
        axis.title=element_text(size=14),
        plot.margin = margin(0, 0, 0, 0))
```

```{r v to v correlations, fig.cap="Correlation of Vertex to Vertex gradients and field surveyed measures using Spearman's correltation coefficient (R).", fig.width=8, fig.height=7, fig.align="left",echo=FALSE}

trimmed.dat %>%
  filter(stream_order < 6)%>%
  ggplot(., aes(x=surveyed_gradient, y=gradient))+
  geom_point()+
  facet_wrap(~stream_order, scales="free")+
  stat_cor(aes(label = after_stat(r.label)), method="spearman")+
  labs(x="Field Surveyed Gradient (%)", 
       y="Vertex-to-Vertex Gradient (%)")+
  theme_classic()+
  theme(axis.text=element_text(size=12),
        strip.text = element_text(size=14),
        axis.title=element_text(size=14))

```

```{r averaged gradient correlations, fig.cap="Correlation of gradient averaged over three vertex to vertext segments with field surveyed gradients using Spearman's correltation coefficient (R).", fig.width=7, fig.height=7, fig.asp=0.9, fig.align="left", echo=FALSE}

trimmed.dat %>%
  filter(stream_order < 6)%>%
  ggplot(., aes(x=surveyed_gradient, y=gradient_avg))+
  geom_point()+
  facet_wrap(~stream_order, scales="free")+
  stat_cor(aes(label = after_stat(r.label)), method="spearman")+
  labs(x="Field Surveyed Gradient (%)", 
       y="Avg V-to-V gradient (nearest, up, and down stream) (%)")+
  theme_classic()+
  theme(axis.text=element_text(size=12),
        strip.text = element_text(size=14),
        axis.title=element_text(size=14))
```

